---
title: "INST 490 COVID-19 Data Visualization"
output: COVID_19_Data_Visualization
---
Across all three graphs, we can see how COVID-19 affected Maryland over time and how the state responded. The first graph highlights that the largest counties, especially Montgomery and Prince George’s consistently reported the most cases. This isn’t surprising given their population sizes, but it shows how much more heavily the pandemic impacted densely populated areas compared to smaller, more rural counties.
The second graph focuses on vaccination rollout by age group. Maryland clearly prioritized older adults first, which is reflected in how quickly the 50–64 and 65+ groups rose in the number of fully vaccinated individuals. These groups hit high vaccination levels early in 2021, while younger age groups didn’t see major increases until later. This pattern matches state and national guidance at the time, which pushed older and higher risk groups to the front of the line.

The final graph brings the trends together. When the vaccination rollout ramped up in early 2021, the growth in statewide COVID-19 cases began to slow. Even though cases continued to rise at different points in the pandemic, the relationship suggests that higher vaccination coverage helped prevent sharper spikes and may have reduced the severity of outbreaks. Overall, these graphs show how COVID-19 spread unevenly across counties, how age played a big role in vaccination timing, and how vaccinations helped stabilize Maryland’s case trends over time.
```{r}
library(tidyverse)

# Load the county-level COVID-19 dataset.
# The file should already be cleaned, so this step simply brings it into R.
county <- read_csv("County.csv")

# Create a line chart showing how daily COVID-19 cases changed over time
# for each Maryland county in the dataset. Colored per county for readability 
# and separation of counties.
ggplot(county, aes(x = Date, y = Cases, color = County)) +
  
  # Plot a separate line for each county. A line graph allows us to document 
  # and separate all the counties on a graph. The slight transparency helps
  # prevent the plot from looking too cluttered when many counties overlap.
  geom_line(alpha = 0.7, linewidth = 0.9) +
  
  # A minimal theme to keep the graphic clean and readable.
  theme_minimal() +
  
  # Labels
  labs(
    title = "Daily COVID-19 Cases in Maryland by County",
    x = "Date",
    y = "Daily Cases",
    color = "County"
  )

```
This graph shows how COVID-10 cases grew across Maryland counties from 2020 onward. Two counties, Montgomery and Prince George's had the highest cumulative case counts throught the pandemic, reflecting their larger populations and early community spread. Mid sized counties like Baltimore, Baltimore City, and Anne Arundel also experienced steady growth, while more rural counties such as Kent, Garrett, and Worcester remained far lower. This graph shows geogrpahic disparity, with urban/suburban counties experiencing far higher burdens than rural areas.

```{r}
# Load the statewide vaccination dataset.
# This file contains cumulative vaccination counts broken down by age group.
state <- read_csv("Statewide.csv")

# Plot the cumulative number of fully vaccinated individuals over time,
# grouped by age range. Each age group gets its own line. Color by age range for
# easy interpretation.
ggplot(state, aes(x = Date, y = FullyVaccinatedCumulative, color = AgeRange)) +
  
  # A line graph here will be easy to look at every age group separately, 
  # especially because there are so few of them.
  geom_line(linewidth = 1.2) +
  
  # Minimal theme to keep the plot simple and easy to read.
  theme_minimal() +
  
  # Labels
  labs(
    title = "Cumulative Fully Vaccinated Maryland Residents by Age Group",
    x = "Date",
    y = "Cumulative Fully Vaccinated",
    color = "Age Group"
  )

```
This graph highlights how vaccination uptake differed dramatically by age group. Older adults, especially the 65+, and 50-64 groups were vaccinated earliest and fastest, rapidly climbing in early 2021 when vaccines first became available. The 18-49 group eventually became the largest vaccinated population due to its size, but with a slower slope. Youth vaccination (12-17, 11 and under) rolled out much later, which explains their lower cumulative totals. The story here is pretty clear, Maryland prioritized older adults, leading to early steep increases in high risk groups, before expanding eligibility to younger populations.

```{r}
# This plot compares statewide COVID-19 case rates with the vaccination rollout.
# Both trends are plotted on the same timeline so we can see how they change together.

ggplot(state, aes(x = Date)) +
  
  # Statewide cases per 100,000 residents.
  # This solid line shows how infection levels changed over time.
  geom_line(aes(y = Cases_per_100), linewidth = 1.1) +
  
  # Cumulative first vaccine doses. A line graph because we're looking at 
  # trends and we can see the change over time easily with a line graph.
  # Since vaccination numbers are much larger, we scale them down (divided by 10) 
  # so both lines can be compared on the same graph.
  # A dashed line helps visually separate vaccination trends from case trends. 
  # A smaller linewidth to see the denser areas in the graph.
  # No color because there are so many cases, we can look at density.
  geom_line(aes(y = FirstDailyDoseCumulative / 10), 
            linewidth = 0.5, 
            linetype = "dashed") +
  
  # Minimal theme to keep the visualization clean and easy to interpret.
  theme_minimal() +
  
  # Labels
  labs(
    title = "Cases per 100K vs. Vaccination Rollout in Maryland",
    x = "Date",
    y = "Cases per 100K (solid) | Vaccination (scaled, dashed)"
  )

```
This combined trend attemps to compare COVID-19 case rates with vaccination rollout over time. Although a bit visually crowded, the major takeaway is that as vaccination coverage increased sharply in early-mid 2021, statewide case rates began to stabilize, and major spiked became less severe compared to early waves. The dashed cavvination line rising alongside a flattening in cases suggests that vaccination helped blunt the spread and severity of COVID-19, even as new variants emerged later on.
